The Goldilocks Level of Friction: Using User Numbers, Ratings, and Reviews to Evaluate Digital Self-Control Tools

### 2022-04-29 paper.Rmd
tweak abstract

---
journal: International Journal of Human-Computer Studies
title: "The Goldilocks Level of Friction: Using User Numbers, Ratings, and Reviews to Evaluate Digital Self-Control Tools"
keywords: Attention, self-control, self-regulation, distraction, digital wellbeing, focus
abstract: |
  Much effort has been invested in designing digital systems that keep people 'hooked', but comparatively little is known about how designers can best support people in re-gaining control over digital device use. Online, hundreds of apps and browser extensions promise to help people self-regulate their use, and their reviews and popularity metrics can indicate which design patterns are found useful in the will. Moreover, they reveal how platforms such as Android and iOS differ in the ecosystems they enable for such tools, which has important practical implications for users struggling with self-regulation difficulties. We analysed user reviews, installation numbers, and ratings for 334 digital self-control tools (DSCTs) on the Google Play, Chrome Web, and Apple App stores, investigating what user reviews reveal about usage contexts and key design challenges, and how functionality relates to popularity metrics. Our thematic analysis of 1,529 reviews (sampled from a data set of 54,320 reviews scraped in March 2019) found that DSCTs are seen as highly important for staying focused on less instantly rewarding tasks when digital distractions are easily available. Users seek DSCTs that adapt to their personal definitions of 'distracting use', and provide incentives sufficient to affect behaviour change without feeling overly intrusive. DSCTs' interventions, even if highly useful, can elicit negative emotions, which is an important reason that the goals of the user and the DSCT should align. Tools that combine two or more types of design patterns (e.g., website blocking and goal reminders) received significantly higher ratings than those implementing a single type, which resonated with feature suggestions in the reviews. We discuss implications for research and design, including how basic psychological research on self-regulation can inform evaluation measures for DSCTs, as well as guide future exploration of how design patterns and framings of 'distraction' interact.

bibliography: references.bib

>>>>

---
journal: International Journal of Human-Computer Studies
title: "The Goldilocks Level of Friction: Using User Numbers, Ratings, and Reviews to Evaluate Digital Self-Control Tools"
keywords: Attention, self-control, self-regulation, distraction, digital wellbeing, focus
abstract: |
  Much effort has been invested in designing digital systems that keep people 'hooked'. By contrast, comparatively little is known about how designers can support people in re-gaining control. Online, however, hundreds of apps and browser extensions promise to help people self-regulate use of digital devices. Reviews and popularity metrics for these digital self-control tools (DSCTs) can indicate which design patterns are useful in the wild. Moreover, they reveal how platforms such as Android and iOS differ in the ecosystems they enable for DSCTs, which has important practical implications for end users. We analysed reviews, installation numbers, and ratings for 334 DSCTs on the Google Play, Chrome Web, and Apple App stores, investigating what user reviews reveal about usage contexts and key design challenges, and how functionality relates to popularity metrics. Our thematic analysis of 1,529 reviews (sampled from a data set of 54,320 reviews scraped in March 2019) found that DSCTs are seen as highly important to stay focused on less instantly rewarding tasks when digital distractions are easily available. Users seek DSCTs that adapt to their personal definitions of 'distracting use', and which provide incentives sufficient to change behaviour without feeling overly intrusive. DSCTs' interventions can elicit negative emotions, which appeared to be an important reason that the DSCT accurately capture users' definition of distraction. Tools that combine two or more types of design patterns (e.g., website blocking and goal reminders) received slightly higher ratings than those implementing a single type. This mirrored suggestions in the reviews to combine different types of design patterns to provide a incentive for behaviour change that is 'just right'. 
  UPDATE FINAL SENTENCE AFTER REWRITE: We discuss implications for research and design, including how basic psychological research on self-regulation can inform evaluation measures for DSCTs, as well as guide future exploration of how design patterns and framings of 'distraction' interact.
bibliography: references.bib
